---
title: Blog 2: The Economy
author: Kaitlyn Vu
date: "2024-09-15"
slug: blog-2
categories: []
tags: []
---

## Introduction
"It's the economy, stupid." Coined by political strategist Jim Carville, the phrase became a hallmark of Bill Clinton's successful election bid against incumbent George H.W. Bush in 1992. Larry Bartels and Christopher Achen note in their 2016 book _Democracy for Realists_ that voters engage in "economic voting," through which economic distress — especially during an election year — harms election outcomes for the incumbent. However, to what extent do Americans truly vote based on the state of the economy? In this blog, I utilize economic factors such as GDP and RDI growth to predict presidential election outcomes and assess the utility of a predictive model based solely on the economy.  

```{r setup, include=FALSE}
# hide code
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

# import libraries
library(tidyverse)
library(ggplot2)
library(car)
library(blogdown)
library(sjPlot)
library(patchwork)
library(knitr)

# import data
d_pv <- read_csv("popvote_1948-2020.csv")
d_fred <- read_csv("fred_econ.csv")
d_bea <- read_csv("bea_econ.csv") |> 
  rename(year = "Year",
         quarter = "Quarter", 
         gdp = "Gross domestic product", 
         gnp = "Gross national product", 
         dpi = "Disposable personal income", 
         consumption = "Personal consumption expenditures", 
         goods = "Goods", 
         durables = "Durable goods", 
         nondurables = "Nondurable goods", 
         services = "Services", 
         pop = "Population (midperiod, thousands)")
```

## Historical Trends with Gross Domestic Product (GDP)
To start, I analyze the relationship between GDP growth and the incumbent party's national popular vote share from past presidential elections. I utilize GDP growth from Quarter 2 (Q2) of an election year: Q2 is the period of April through June, which provides a stable but recent snapshot of the pre-election economy.  

Notably, the plot on the left includes all the presidential elections from 1948-2020, while the plot on the right excludes 2020. The 2020 contest between President Joe Biden and former President Donald Trump is a significant outlier: unprecedented economic turmoil stemming largely from the COVID-19 pandemic [tanked](https://www.reuters.com/markets/us/us-election-uncertainty-slowdown-heady-mix-markets-mcgeever-2024-07-03/) the national economy during that election year.  

```{r historical gdp, fig.width=12, fig.height=6}
# Q2 GDP (all)
d_inc_econ <- d_pv |> 
  filter(incumbent_party == TRUE) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))

reg_gdp <- lm(pv2p ~ GDP_growth_quarterly, 
               data = d_inc_econ)

gdp_plot_1 <- d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "1948-2020 (Y = 51.26 + 0.27 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14),
        plot.margin = margin(10, 30, 0, 0))

# Q2 GDP (without 2020)
d_inc_econ_2 <- d_inc_econ |>
  filter(year != 2020)

reg_gdp_2 <- lm(pv2p ~ GDP_growth_quarterly, 
                         data = d_inc_econ_2)

gdp_plot_2 <- d_inc_econ_2 |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "1948-2016 (Y = 49.38 + 0.74 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 0, 0, 30))

# plot 1: Q2 GDP and national popular vote share (I used ChatGPT to help format the plots side-by-side)
gdp_plot_1 + gdp_plot_2 + 
  plot_annotation(title = "Incumbent Party's National Popular Vote Share and Q2 GDP Growth",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```

This is a regression table for the incumbent party's national popular vote share and Q2 GDP growth for the presidential elections from 1948-2016. 

```{r historical gdp reg table}
# table 1: Q2 GDP and national popular vote share
tab_model(reg_gdp_2, show.se = TRUE, 
          title = "Regression Table for Q2 GDP Growth (1948-2016)", 
          dv.labels = "Incumbent Party's National Popular Vote Share")
```

As expected, there is a positive relationship between Q2 GDP growth and the incumbent party's national popular vote in past presidential elections. This correlation strengthens when we exclude the 2020 election: the slope of the line of best fit increases from 0.27 to 0.74. However, the R-squared value of approximately 0.2 for the 1948-2016 plot suggest that this relationship isn't particularly strong. R-squared values measure how much variation in the dependent variable (incumbent party's national vote share in this case) is captured by the fitted model's predicted value — offering an ease of interpretability for in-sample fit. However, it is important to note that utilizing in-sample model fit alone can contribute to overconfidence or over-fitting. 

## Historical Trends with Real Disposable Personal Income (RDI)
Similarly, I examine the relationship between Q2 RDI growth and the incumbent party's national popular vote share. I include RDI in this blog to engage with more individualistic perceptions of voter behavior like "pocketbook voting": voters may prioritize their own economic well-being over broader national  indicators like GDP. The left plot includes presidential elections from 1948-2020, while the right plot excludes 2020.

```{r historical rdi, fig.width=12, fig.height=6}
# Q2 RDI (all)
reg_rdi <- lm(pv2p ~ RDPI_growth_quarterly, 
               data = d_inc_econ)

rdi_plot_1 <- d_inc_econ |> 
  ggplot(aes(x = RDPI_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter RDI Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "1948-2020 (Y = 51.97 - 0.03 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 25, 0, 0))

# Q2 RDI (without 2020)
reg_rdi_2 <- lm(pv2p ~ RDPI_growth_quarterly, 
               data = d_inc_econ_2)

rdi_plot_2 <- d_inc_econ_2 |> 
  ggplot(aes(x = RDPI_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter RDI Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "1948-2016 (Y = 49.87 + 0.46 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
       plot.margin = margin(10, 0, 0, 25) )

# plot 2: Q2 RDI and national popular vote share 
rdi_plot_1 + rdi_plot_2 + 
  plot_annotation(title = "Incumbent Party's National Popular Vote Share and Q2 RDI Growth",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```

This is a regression table for the relationship between the incumbent party's national popular vote share and Q2 RDI growth for the presidential elections from 1948-2016. 

```{r historical rdi table}
# table 2: Q2 RDI and national popular vote share
tab_model(reg_rdi_2, show.se = TRUE, 
          title = "Regression Table for Q2 RDI Growth (1948-2016)", 
          dv.labels = "Incumbent Party's National Popular Vote Share")
```

With RDI, excluding 2020 makes quite a significant difference: the slope of the best fit line jumps from -0.03 to 0.46. Again, there appears to be a positive correlation between Q2 RDI growth and the incumbent party's national popular vote share in past presidential elections. The R-squared values here are lower than those for GDP growth, however, which suggests that the Q2 RDI growth is also not a strong predictor of the incumbent party's national popular vote share. 

## Sitting Presidents vs Same-Party Heirs
Next, I examine whether the effect of the economy is stronger for sitting presidents running for re-election compared to same-party heirs seeking the office. First, I sorted the data into elections where the incumbent was running for re-election (11 elections) and elections where a candidate of the incumbent's party was running (7 elections). The left plot is for sitting presidents, while the right plot is for same-party heirs. Once again, I exclude 2020 from the analysis.

```{r comparison gdp, fig.width=12, fig.height=6}
# sitting Q2 GDP (all)
d_sitting_econ <- d_pv |> 
  filter(incumbent == TRUE, 
         year != 2020) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))

reg_sitting_gdp <- lm(pv2p ~ GDP_growth_quarterly, 
               data = d_sitting_econ)

sitting_plot_gdp <- d_sitting_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "National Popular Vote Share", 
       title = "Sitting Presidents (Y = 50.43 + 0.85 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 25, 0, 0))

# heir Q2 GDP (all)
d_heir_econ <- d_pv |> 
  filter(incumbent_party == TRUE &
         incumbent == FALSE) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))

reg_heir_gdp <- lm(pv2p ~ GDP_growth_quarterly, 
               data = d_heir_econ)

heir_plot_gdp <- d_heir_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "National Popular Vote Share", 
       title = "Same-Party Heirs (Y = 48.46 + 0.29 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 0, 0, 25))

# plot 3: sitting vs heir gdp
sitting_plot_gdp + heir_plot_gdp + 
  plot_annotation(title = "Sitting Presidents vs Same-Party Heirs for Q2 GDP Growth",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```

The associated regression table is below.

```{r comparison gdp table}
# table 3: sitting vs heir gdp
tab_model(reg_sitting_gdp, reg_heir_gdp, show.se = TRUE, 
          title = "Comparison Regression Table for Q2 GDP Growth (1948-2020)", 
          dv.labels = c("Sitting Presidents", "Same-Party Heirs"))
```

With a slope of 0.85, it seems that there is a stronger correlation between Q2 GDP growth and national popular vote share for sitting presidents running for re-election than same-party heirs (0.29). The R-squared value for the sitting presidents model is also larger (0.44 vs 0.11), which means that Q2 GDP growth is a much stronger predictor of the national popular vote share for sitting presidents than same-party heirs.

I created the same visualizations for Q2 RDI growth. 

```{r comparison rdi, fig.width=12, fig.height=6}
# sitting Q2 RDI (all)
reg_sitting_rdi <- lm(pv2p ~ RDPI_growth_quarterly, 
               data = d_sitting_econ)

sitting_plot_rdi <- d_sitting_econ |> 
  ggplot(aes(x = RDPI_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter RDI Growth (%)", 
       y = "National Popular Vote Share", 
       title = "Sitting Presidents (Y = 50.39 + 0.69 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 25, 0, 0))

# heir Q2 RDI (all)
reg_heir_rdi <- lm(pv2p ~ RDPI_growth_quarterly, 
               data = d_heir_econ)

heir_plot_rdi <- d_heir_econ |> 
  ggplot(aes(x = RDPI_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter RDI Growth (%)", 
       y = "National Popular Vote Share", 
       title = "Same-Party Heirs (Y = 51.49 - 0.46 * X)") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 14), 
        plot.margin = margin(10, 0, 0, 25))

# plot 4: sitting vs heir rdi
sitting_plot_rdi + heir_plot_rdi + 
  plot_annotation(title = "Sitting Presidents vs Same-Party Heirs for Q2 RDI Growth",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```


```{r comparison rdi table}
# table 4: sitting vs heir rdi
tab_model(reg_sitting_rdi, reg_heir_rdi, show.se = TRUE, 
          title = "Comparison Regression Table for Q2 RDI Growth (1948-2020)", 
          dv.labels = c("Sitting Presidents", "Same-Party Heirs"))
```

For RDI, sitting presidents have historically enjoyed a positive correlation between Q2 RDI growth and national popular vote share with a slope of 0.69 for the line of best fit. On the other hand, there appears to be a negative relationship between Q2 RDI growth and national popular vote share for same-party heirs with a slope of -0.46. Again, the R-squared value for the sitting presidents model is a bit greater (0.28 vs 0.21), which means that Q2 RDI growth is a stronger predictor of the national popular vote share for sitting presidents than same-party heirs. 

## Predictions for 2024
While such univariate models are imperfect, we can still use them to develop predictions for the 2024 election. The following table utilizes the regression models included in this blog to estimate the national popular vote share for the incumbent party's candidate; in this case, Vice President Kamala Harris is a same-party heir so those equations also apply. Based on the table, Harris is predicted to receive a bit over 50% of the national popular vote this November.

```{r predictions}
# Q2 GDP and RDI for 2024
GDP_2024 <- d_fred |> 
  filter(year == 2024 & quarter == 2) |> 
  select(GDP_growth_quarterly)

RDI_2024 <- d_fred |> 
  filter(year == 2024 & quarter == 2) |> 
  select(RDPI_growth_quarterly)

# predict 2024 national popular vote share
gdp <- predict(reg_gdp_2, GDP_2024, interval = "prediction")
rdi <- predict(reg_rdi_2, RDI_2024, interval = "prediction")
heir_gdp <- predict(reg_heir_gdp, GDP_2024, interval = "prediction")
heir_rdi <- predict(reg_heir_rdi, RDI_2024, interval = "prediction")

gdp <- round(gdp, 2)
rdi <- round(rdi, 2)
heir_gdp <- round(heir_gdp, 2)
heir_rdi <- round(heir_rdi, 2)

# join into 1 data frame (I used ChatGPT to create this data frame)
predictions <- data.frame(
  Model = c("GDP", "Heir GDP", "RDI", "Heir RDI"),
  Fit = c(gdp[, "fit"][1], heir_gdp[, "fit"][1], rdi[, "fit"][1], heir_rdi[, "fit"][1]),
  Lwr = c(gdp[, "lwr"][1], heir_gdp[, "lwr"][1], rdi[, "lwr"][1], heir_rdi[, "lwr"][1]),
  Upr = c(gdp[, "upr"][1], heir_gdp[, "upr"][1], rdi[, "upr"][1], heir_rdi[, "upr"][1])
)

# table 5: 2024 predictions
knitr::kable(predictions, 
             col.names = c("Model", "Fit", "Lower CI", "Upper CI"))
```

## Conclusion
In very general terms, it seems that the incumbent's party enjoys an advantage when the economy is doing well. Furthermore, the effect of the economy appears to be stronger for sitting presidents running for re-election than same-party heirs. 

However, the relatively low R-squared values of all these simple models suggest that GDP and RDI growth only loosely fit national popular vote data. As such, continuing to add different variables — beyond the economy — to my model can help to increase its predictive performance. The predictive power of the economy can vary over time due to factors such as unexpected economic shocks (e.g. COVID-19), other policy concerns, and/or increasing political polarization.
